---
title: "Using `dogesr` to find out about the doges families"
author: "JJ Merelo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using `dogesr` to find out about the doges families}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/doges.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Only men belonging to one of the Venetian noble families could actually become doge after the *Serrata*; however, even if the number of noble families was initially small, it was eventually expanded to include several hundred noble families [@lane2019enlargement]. Not all of them, however, were able to include one of their own in the list of doges; on the other hand, the same serrata brought about informal mechanisms that guaranteed a fast turnover in the job of doge, so that many families would get the chance [@histories:jj]

Using data from `dogesr` [@dogesr], we will, in this vignette, have a look at these families, who they were, and how many of them were there.

## Set up

We load the dataset needed, called `doge.families`.

```{r load}
devtools::load_all(".")
data("doge.families")
```

This will import the data from the `dogesr` package into the `doge.families` *tibble*.


## Conclusions

## References
