---
title: "Using `dogesr` to work with the social network of Venetian doges' families"
author: "JJ Merelo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using `dogesr` to work with the social network of Venetian doges' families}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/doges.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Using data from `dogesr` [@dogesr], we will, in this vignette, analyze a sample of the social network of Venetian noble families, the one that includes *only* marriages by nobles that became doges and, in some cases, their parents.

## Set up

```{r load, warning=FALSE,message=FALSE}
library(dogesr)
library(igraph)
doges.sn <- marriage.graph()
```

This will call the function that creates, from raw data, the `igraph` data structure with which we can work. We still need to simplify this net to be able to work with it. 
```{r simplify}
E(doges.sn)$weight <- 1
doges.sn <- simplify(doges.sn, edge.attr.comb=list(weight="sum"))
V(doges.sn)$degree <- degree(doges.sn)
```

## Conclusions

## References
