---
title: "Ruling for life: Life and Times of Venetian doges"
author: "JJ Merelo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ruling for life: Life and Times of Venetian doges}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/doges.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Since doges were elected for life, it was apparently a conscious design by the electoral college to choose them in such a way that their terms would not last too long [@smith2021long]. Using data from `dogesr` [@dogesr], we will, in this vignette, see to what extent that happened and if there was some evolution during the time the Republic of Venice existed

## Set up

```{r load}
library(dogesr)
data("doges")
```

This will import the data from the `dogesr` package into the `data.doges` data frame. We are interested in the `Doges` and `Years` variables. Let's create a data frame out of that:

```{r years}
doges.years <- data.frame( doge = data.doges$Doge, years = data.doges$Years)
```


## References
