---
title: "Using `dogesr` to work with Venetian doges family types"
author: "JJ Merelo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using `dogesr` to work with Venetian doges family types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/doges.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

For the longest part of its existence, Venetian noble families used tactical marriage to improve the lot of their *case* or dynasties. Since access to nobility was closed by the beginning of the XIV century, and then enlarged little by little, there was a kind of stratification in the nobility, and marrying *older* families was a way of improving the political, commercial and social prospects for your own family.

Using data from `dogesr` [@dogesr], we will, in this vignette, see to what extent that happened and how different types of families married with each other.

## Set up

We load the datasets needed; `doges` describes doges and their marriages, `families` includes data for every noble family.

```{r load}
devtools::load_all(".")
data("doges")
data("families")
```

This will import the data from the `dogesr` package into the `data.doges` data frame, and family data in the `family.types` list. Let's merge them into a single data frame.

```{r years}
marriages.types <- 
  data.frame( doge.family = data.doges$Family.doge,
              dogaressa.family = data.doges$Family.dogaressa,
              fam.type.doge = unname( unlist ( family.types[ data.doges$Family.doge ] )),
              fam.type.dogaressa = unname( unlist ( family.types[ data.doges$Family.dogaressa ] ))
)
                               
```

## Conclusions


## References
