---
title: "Using `dogesr` to work with Venetian doges family types"
author: "JJ Merelo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using `dogesr` to work with Venetian doges family types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/doges.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

For the longest part of its existence, Venetian noble families used tactical marriage to improve the lot of their *case* or dynasties. Since access to nobility was closed by the beginning of the XIV century, and then enlarged little by little, there was a kind of stratification in the nobility, and marrying *older* families was a way of improving the political, commercial and social prospects for your own family.

Using data from `dogesr` [@dogesr], we will, in this vignette, see to what extent that happened and how different types of families married with each other.

## Set up

We load the datasets needed; `doges` describes doges and their marriages, `families` includes data for every noble family.

```{r load}
devtools::load_all(".")
data("doges")
data("families")
```

This will import the data from the `dogesr` package into the `data.doges` data frame, and family data in the `family.types` list. Let's merge them into a single data frame. The table of all family names and their corresponding family type is shown below.`

```{r years}
marriages.types <- 
  data.frame( doge.family = data.doges$Family.doge,
              dogaressa.family = data.doges$Family.dogaressa,
              fam.type.doge = unname( unlist ( family.types[ data.doges$Family.doge ] )),
              fam.type.dogaressa = unname( unlist ( family.types[ data.doges$Family.dogaressa ] ))
)
knitr::kable(marriages.types)
```

Let's create an adjacency matrix for this

```{r types}
marriages.just.types <- marriages.types
marriages.just.types$doge.family <- NULL
marriages.just.types$dogaressa.family <- NULL
marriages.just.types$fam.type.doge <- paste0(marriages.just.types$fam.type.doge,"♂")
marriages.just.types$fam.type.dogaressa <- paste0(marriages.just.types$fam.type.dogaressa,"♀")
levs <- c(unique(unlist(marriages.just.types$fam.type.doge, use.names = FALSE)),unique(unlist(marriages.just.types$fam.type.dogaressa, use.names = FALSE)))
types.adjacency <- table(lapply(marriages.just.types, factor, levs))
types.adjacency <- types.adjacency[,-c(1:9)]
types.adjacency <- types.adjacency[-c(10:18),]
knitr::kable(types.adjacency)
```

The *heaviest* link seem to be between the *evangeliche* and *apostoliche* *cases*, although "Not available" is the heaviest column for doges, who either did not marry or married some non-patrician family; this happened most frequently in the pre-Serrata times.

Let's look at this in a Sankey diagram

```{r sankey}
library(tidyr)
library(dplyr)
library(tibble)
links <- types.adjacency  %>% as.data.frame() 
nodes <- data.frame(
  name=c(as.character(links$fam.type.doge), as.character(links$fam.type.dogaressa)) %>% 
    unique()
  )
links$IDsource <- match(links$fam.type.doge, nodes$name)-1 
links$IDtarget <- match(links$fam.type.dogaressa, nodes$name)-1

library(networkD3)
links <- links[ links$Freq > 0,]
sankeyNetwork(Links = links, Nodes = nodes,
                     Source = "IDsource", Target = "IDtarget",
                     Value = "Freq", NodeID = "name", 
                     sinksRight=FALSE)
```

The "NA" category of families seems to dominate the female part of the diagram, corresponding to dogaresse. We are mostly interested, however, in the social links formed by different types of families, so we will simply eliminate those rows and columns to have a clearer picture.

```{r non.na}
types.adjacency <- types.adjacency[-1,-1]
links2 <- types.adjacency  %>% as.data.frame() 
nodes2 <- data.frame(
  name=c(as.character(links2$fam.type.doge), as.character(links2$fam.type.dogaressa)) %>% 
    unique()
  )
links2$IDsource <- match(links2$fam.type.doge, nodes2$name)-1 
links2$IDtarget <- match(links2$fam.type.dogaressa, nodes2$name)-1
links2 <- links2[ links2$Freq > 0,]
sankeyNetwork(Links = links2, Nodes = nodes2,
                     Source = "IDsource", Target = "IDtarget",
                     Value = "Freq", NodeID = "name")
```

## Conclusions


## References
