---
title: "It's a doges' life: examining term limits in Venetian doges life tenure"
author: "JJ Merelo"
date: "`r Sys.Date()`"
output:
  pdf_document:
    keep_tex: true
bibliography: ../inst/doges.bib
abstract: |
  During most of the lifespan of the Venetian republic, doges or presidents were elected for life. However, a long tenure was a rare event, which effectively resulted in term limits, as found by [@smith2021long]. In this paper we examine these tenures and its evolution during the existence of the Venetian republic, trying to prove Smith et al. thesis, but also the event or events that effectively caused that limitation by design. Finally, we will discuss for the causes of this limitation and its effective consequences.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dogesr)
data("doges")
doges.all.years <- data.frame( doge = data.doges$Doge, years = data.doges$Years)

```

## Introduction

The Venetian republic, which existed from the seventh to the end of the eighteenth century, was a unique polity in many different aspects, including the fact that it existed for more than one thousand years essentially under the same set of laws and its stability in a geographic milieu, the Italian peninsula, that underwent many waves of upheaval under the opposite forces of the Papal states, the Holy Roman Empire, and later the Ottoman, French, Spanish and Austro-Hungarian empires [@madden2012venice].

Many explanations have been sought for this fact, and of course the governing institutions have been examined, including the head of government, the so called *doge* (*doxe* in the vernacular Veneto language; the term is equivalent to "duke" approximately in other languages). Doges were elected for life by an electoral college composed of 41 nobles, which were also elected through a series of lotteries and nominations emanating from the *Maggior Consiglio*, which included all noble families in Venice [@molinari2020republic].

Although initially this electoral council was composed of all persons with a certain economic position or at any rate chosen by the doge, by the beginning of the XIV century  a *closure* (*serrata*) took place [@serrata]: seats were closed to only a few families, hereafter called *vecchie* or "old" families.

This had a series of consequences, including the creation of a (relatively) egalitarian society [@Stockwell2011], at least among the nobles; according to other authors [@serrata], this had the opposite effect of creating a chasm between the aristocrats and the *popolani*, or common folk, which ultimately brought the demise of the Republic.

Be that as it may, this single fact had to be compounded with others to account for the fact that the system persisted through several centuries, and that except for a single case, no doge had to be indicted of corruption or treason, or for that matter simply destituted or murdered, as was the case in many of the other Italian city-states or Europe at large.

Another factor might be at play here. Since doges were elected for life, it was apparently a conscious design by the electoral college to choose them in such a way that their terms would not last too long [@smith2021long]. Using data from `dogesr` [@dogesr], we will, in this paper, see to what extent that happened and if there was some evolution during the time the Republic of Venice existed.

The rest of the paper is organized as follows. Next we will show the state of the art in studies of how doge election and related choices affected the stability of the Venetian republic. Then we will analyze these terms, and its evolution with time, check if there were some shifts in the tenure duration. Finally, we will present our conclusions and discuss its implications.

## State of the art


## Examining doges' terms

There were a number of doges whose tenure spanned one or two natural years, as is shown below:

```{r one.year.or.less, echo = FALSE}
doges.one.year <- unique(doges.all.years[doges.all.years$years <= 1,])
knitr::kable(doges.one.year[order(doges.one.year$years),],col.names=NULL,caption="Doges whose tenure was within a natural year (indicated with 0 in the last column) or went from one year to the next. The first column is the order or accession to dogarate.")
```

So it effectively looks like most doges effectively had a short shelf life. As a matter of fact, the doge that staid in power the longest was

```{r longest.doge, echo = FALSE}
knitr::kable(doges.all.years[which.max(doges.all.years$years),])
```

Francesco Foscari was elected when he was 50, defeating Pietro Loredan. After being in power for 34 years, he was the only doge forced to abdicate after the *serrata* or new rules electing the doges.

We can see whether this has any kind of influence using now several functions provided by the `dogesr` package next.

## Analyzing the evolution of the effective terms of the doges using `doges.years`

This can be analyzed a but further with a function provided by `dogesr`, `doges.years`. This filters part of the original data, leaving only data needed to work with the doges' terms (or years)

```{r using.doges.years, echo = FALSE}
doges.years.all <- doges.years()
doges.years.all$Century <- doges.years.all$Century +1 
knitr::kable(head(doges.years.all[order(-doges.years.all$Years),],10))
```

First thing we can note is that, among the top 10, there is only one, Leonardo Loredan, who was elected after the aforementioned Francesco Foscari. This was, according to everyone, an unexpected event [@loredan], and he was chosen by the minimum needed to get elected. So he was really an outlier. The distribution in bins of 5 years, below, shows how infrequent was a term of more than 10 years.

```{r distribution, fig.height=3, fig.pos="!H", echo = FALSE, fig.cap="This histogram reproduces, with a fixed bin size, Figure 1 in [@smith2021long]."}
library(ggplot2)
ggplot(doges.years.all, aes(x=Years))+geom_histogram(binwidth=5)
```

The distribution is skewed towards very short terms because it includes the first years of the institution, when murder or destitution by popular revolt was relatively usual; it also shows a relatively long tail exactly for the same reasons. This is the distribution by centuries

```{r boxplot, echo = FALSE, fig.pos="!H",}
doges.years.all$Century <- as.factor(doges.years.all$Century)
ggplot(doges.years.all, aes(x=Century,y=Years))+geom_boxplot()
```

There seems to be a clear difference before and after the thirteenth century; there is also a return to longer tenures by the end of the Republic, in the 18th century. Was there a real difference?

```{r serrata, echo = FALSE, fig.pos="!H",}
doges.years.all$pre.serrata <- TRUE
doges.years.all[doges.years.all$Start >= 1297,]$pre.serrata <- FALSE
means <- aggregate(Years ~  pre.serrata, doges.years.all, mean)
medians <- aggregate(Years ~  pre.serrata, doges.years.all, median)
ggplot(doges.years.all, aes(x=pre.serrata, y=Years))+geom_boxplot(notch=T)+
  stat_summary(fun=mean, geom="point", shape=20, size=3, color="red", fill="red") +
  geom_text(data = means, aes(label = round(Years, 2), y = Years + 2), size = 3) + 
  geom_text(data = medians, aes(label = round(Years, 2), y = Years - 1), size = 3)
```

This difference is significant, as indicated by the Wilcoxon test

```{r wilcox, echo = FALSE}
wtest <- wilcox.test(doges.years.all[doges.years.all$pre.serrata == T,]$Years,   doges.years.all[doges.years.all$pre.serrata == F,]$Years )
```

We can check this hypothesis via change-point detection, an algorithm that checks a time series and points to the precise moment where the change occurred.

```{r change.point, echo=FALSE}
library(trend)
cp.estimate <- lanzante.test(doges.all.years$years)
change.doge <- doges.all.years[as.integer(cp.estimate$estimate),]$doge
```

Using Lanzante's test [@lanzante] from the `trend` R package, this change occurs in the moment `r change.doge` took office.


## Conclusions

As indicated by [@smith2021long], there seems to be evidence in the term limits of doges of the Republic of Venice. In their paper they indicate 1172 as a departure point for the term "limits", when the Maggior Consiglio was formed. According to the figure above, there does not seem to be a significant difference between the XII and XIII century in terms of tenure. However, 1297 does seem to be a watershed event, with all terms after that being significantly different to terms prior to it. We can conclude, then, that the Serrata, closing of the Maggior Consiglio to all but a few patrician families, was in fact the event that marked the shift to electing older doges.


## Acknowledgments

This paper has been supported in part by project and DemocratAI PID2020-115570GB-C22.

## References
